---
title: "Recent trends in cabinet size"
subtitle: "Average number of parties in government peaked in the last decade"
authors:
- admin
date: "2021-04-20"
output: html_document
---

Many countries are governed by not one, but multiple political parties at the same time. Indeed, number of parties in government --- cabinet size --- is rarely one. 

In this blogpost, I explore the average cabinet size, and how this average has been trending in the recent history.

To do so, I turn to the [Parliaments and Governments Database](http://www.parlgov.org/) (ParlGov), which brings together three inter-related datasets listed below. Here I work with only one of them: *Cabinet*.


| Dataset   | Filename      | Observations |
|-----------|---------------|--------------|
| Parties   | view_party    | 1700         |
| Elections | view_election | 1000         |
| Cabinets  | view_cabinet  | 1600         |

To start with, let's load three packages that I need for this task, `dataverse` (to download the dataset from [Harvard Dataverse](https://dataverse.harvard.edu/)),  `tidyverse` (to clean, tidy, and plot the data), and `stargazer` (to table the data).

```{r, load, warning=FALSE, message=FALSE, tidy=FALSE}

# load the packages needed

library(dataverse)
library(tidyverse)
library(stargazer)

```

We can now download the data with the `get_dataframe_by_name` function. 

```{r, download, warning=FALSE, message=FALSE, tidy=FALSE}

# download the data from harvard dataverse

df_cabs <- get_dataframe_by_name(
  filename  = "view_cabinet.tab",
  dataset   = "10.7910/DVN/Q6CVHX",
  server    = "dataverse.harvard.edu"
  )

```

As of `r format(Sys.time(), '%d %B %Y')`, there are `r nrow(df_cabs)` observations and `r length(df_cabs)` variables in the dataset. For the numerical variables, we can create a table with summary statistics with the `stargazer` function.

```{r, summarise, results="asis", warning=FALSE, message=FALSE, tidy=FALSE}

# download the dataset from harvard dataverse

stargazer(as.data.frame(df_cabs), 
          summary = TRUE, type = "html",
          summary.stat = c("mean", "sd", "min", "max"),
          digits = 0.1)

```

Before we can visualise the trend over the years, we first need to get the data into the right shape. There are many helpful functions in the `tidyverse` family for this purpose.

```{r, clean,  warning=FALSE, message=FALSE, tidy=FALSE}

# tidy the data

df_ends <- df_cabs %>% 
        filter(start_date > as.Date("1948-12-31") & start_date < as.Date("2020-01-01")) %>%
        mutate(end_year = as.numeric(format(start_date,'%Y'))) %>%
        select(previous_cabinet_id, end_year) %>%
        distinct()

df <- df_cabs %>%
      filter(start_date > as.Date("1948-12-31") & start_date < as.Date("2020-01-01"))  %>%
      mutate(start_year = as.numeric(format(start_date,'%Y'))) %>%
      group_by(cabinet_id, start_year) %>%
      summarise(parties = sum(cabinet_party)) %>%
      ungroup() %>%
      left_join(., df_ends, by = c("cabinet_id" = "previous_cabinet_id")) %>%
      mutate(end_year = replace_na(end_year, 2019)) %>%
      pivot_longer(cols = c("start_year", "end_year"), values_to = "year") %>%
      select(-name) %>%
      group_by(cabinet_id) %>%
      complete(cabinet_id, year = full_seq(year, 1)) %>%
      fill(parties) %>%
      group_by(year) %>%
      summarise(cabinet_size = mean(parties))

```

Now that the data is in the right shape, Figure \@ref(fig:plot) plots it.

```{r, plot, fig.cap='Average cabinet size, 1949--2019.', warning=FALSE, message=FALSE, tidy=FALSE}

# plot the data

ggplot(df, aes(x = year, y = cabinet_size)) +
        geom_line(size = 1.5) +
        theme_minimal() +
        theme(axis.title = element_text(size = 14),
              axis.text = element_text(size = 14)) +
        labs(y = "Average cabinet size\n", x = "")

```

The figure shows that the average number of parties in government peaked in the last decade.
